---
title: "Univariate_baci_exploration_2020"
author: "Lia Domke"
date: "2/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Initial set up of the first chapter of the master thesis was to look at the potential impact of sea otters on fish communities and their composition. Through using historical seines done by NOAA between 1998-2011 BEFORE sea otters came into areas in Prince of Wales I can complete a BACI (before after controlled impact) study. 

Existing data: *before* seines throughout Prince of Wales and Southeast alaska in 4 habitat types (only 2 of interest--seagrass and understory kelp) done between 1998-2011 AND *after* seines from Prince of Wales in 2017 and 2019. 

These data do not include the control treatement of the experiment, only the impact. 

Steps:

1. include libraries and read in data

2. calculate difference between counts

3. visualize abundance differences

# 1. libraries/data
libraries
```{r include=FALSE}
library(tidyverse)
library(forcats)
```

data
```{r include=FALSE}
ba <- read.csv("../Data/Before_After_Combined_2-2020.csv", stringsAsFactors = FALSE, header = TRUE)
```

required functions 
```{r}
bf.ttest <- function(e,x,cutoff){
#Get test data and sorting data list names
dname<-deparse(substitute(e))
xname<-deparse(substitute(x))
#Partition list into two groups
e1<-cbind(x,e)[x<=cutoff,2]; e2<-cbind(x,e)[x>cutoff,2]
n<-length(x); n1<-length(e1); n2<-length(e2)
#Obtain group medians
e1.med<-median(e1); e2.med<-median(e2)
#Obtain deviations from medians
d1<-abs(e1-e1.med); d2<-abs(e2-e2.med)
#Obtain means of deviations from medians
d1.ave<-mean(d1); d2.ave<-mean(d2)
#Compute the pooled variance of deviations from the medians
p.var<-(sum((d1-d1.ave)^2)+
                  sum((d2-d2.ave)^2))/(n-2)
#Calculate the test statistic
tbf<-abs(d1.ave-d2.ave)/sqrt(p.var/n1+p.var/n2)
#Compute the p-value using the t-distribution
pv<-round(2*pt(-tbf,n-2),4)
#Store the results in a list
results<-list(statistic=c(t=tbf),parameters=c(df=n-2),
        p.value=pv, method=paste("Brown-Forsyth test ranked by",
          xname," and split at ",xname,"=",cutoff),data.name=dname,
                              alternative=c("Variances are unequal"))
#Format results as class hypothesis test and output
class(results)<-"htest";return(results)}

```

# 2. Calculate difference
## 2.1 Pair sites
```{r include=FALSE}
head(ba)
ba$abundance <- as.numeric(ba$abundance)

# sum by site and seine
fish.cpue <- ba %>%
  group_by(siteID_NOAA, date, sp_code, before_after, habitat) %>%
  dplyr::summarise(CPUE = sum(abundance)) %>%
  unite("site_date", c("siteID_NOAA", "date"), remove = FALSE)

fish.cpue$CPUE <- as.numeric(fish.cpue$CPUE)
# change data structure
 
ba_wide <- pivot_wider(fish.cpue, names_from = sp_code, values_from = CPUE)
ba_wide[is.na(ba_wide)] <- 0

# look at all sites
df1 <- ba %>%
  dplyr::select(c(siteID_NOAA, before_after, date, habitat)) %>%
  distinct()

# choose paired sites that were sampled closest to each other
site_sub <- df1[c(1:14,18:21,25,27,32,36:39,40,42,44,46,48,53,56),]
site_sub <- unite(site_sub, "site_date", c("siteID_NOAA", "date"))

# subset data by the sites
ba_pair <- subset(ba_wide, site_date %in% site_sub$site_date)

ba_pair[15,5] <- "kelp"

ba_long <- pivot_longer(ba_pair, cols = -c(site_date, siteID_NOAA, date, 
                                             before_after,habitat), 
                          names_to = "sp_code", values_to = "CPUE")
```

## 2.2 create diff column
```{r include=FALSE}
before <- ba_long %>%
  ungroup(date) %>%
  filter(before_after == "before") %>%
  mutate(CPUE_before = CPUE) %>%
  dplyr::select(-c(CPUE, site_date, date, before_after))

after <- ba_long %>%
  ungroup(date) %>%
  filter(before_after == "after") %>%
  mutate(CPUE_after = CPUE) %>%
  dplyr::select(-c(CPUE, site_date, date, before_after))

ba_long2 <- left_join(before, after, by = c("siteID_NOAA", "sp_code", "habitat"))

ba_diff <- ba_long2 %>%
  group_by(siteID_NOAA, sp_code) %>%
  dplyr::mutate(diff = CPUE_before - CPUE_after)

```

# 3. visualize difference

```{r}
# all sites 

eel <- ba_diff %>%
  group_by(sp_code) %>%
  filter(habitat == "eelgrass") %>%
  filter(diff < 0 | diff > 0) %>%
  ggplot() + geom_point(aes(x = sp_code, y = diff, colour = sp_code))

eel + facet_wrap(~siteID_NOAA)

kelp <- ba_diff %>%
  group_by(sp_code) %>%
  filter(habitat == "kelp") %>%
  filter(diff < 0 | diff > 0) %>%
  ggplot() + geom_point(aes(x = sp_code, y = diff, colour = sp_code))

kelp + facet_wrap(~siteID_NOAA)

ba_diff %>%
  group_by(sp_code) %>%
  mutate(avg_cpue_diff = mean(diff), sd_cpue_diff = sd(diff)) %>%
  ggplot() +
  geom_point(mapping = aes(x = sp_code, y = avg_cpue_diff)) +
  geom_errorbar(aes(x = sp_code, ymin = avg_cpue_diff - sd_cpue_diff, ymax = avg_cpue_diff + sd_cpue_diff), width = .1) +
  xlab("Species code") + ylab("Average difference in cpue") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 24))

# average by before/after by habitat
eel_diff <- ba_diff %>%
  filter(habitat == "eelgrass") %>%
  group_by(sp_code) %>%
  mutate(avg_cpue_diff = mean(diff), sd_cpue_diff = sd(diff)) %>%
  filter(avg_cpue_diff < -5 | avg_cpue_diff > 5) %>%
  ggplot() +
  geom_point(mapping = aes(x = sp_code, y = avg_cpue_diff)) +
  geom_errorbar(aes(x = sp_code, ymin = avg_cpue_diff - sd_cpue_diff, ymax = avg_cpue_diff + sd_cpue_diff),
                width = .1) +
  xlab("Species code") + ylab("Average difference in CPUE ± s.d.") +
  ggtitle("Eelgrass communities") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 24))
  

kelp_diff <- ba_diff %>%
  filter(habitat == "kelp") %>%
  group_by(sp_code) %>%
  mutate(avg_cpue_diff = mean(diff), sd_cpue_diff = sd(diff)) %>%
  filter(avg_cpue_diff < -5 | avg_cpue_diff > 5) %>%
  ggplot() +
  geom_point(mapping = aes(x = sp_code, y = avg_cpue_diff)) +
  geom_errorbar(aes(x = sp_code, ymin = avg_cpue_diff - sd_cpue_diff, ymax = avg_cpue_diff + sd_cpue_diff),
                width = .1) +
  xlab("Species code") + ylab("Average difference in CPUE ± s.d.") +
  ggtitle("Understory kelp communities") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 24))

library(patchwork)
eel_diff + kelp_diff
```

## 3.1. Statistics
```{r}
# divide by hab
long_eel <- ba_long %>%
  filter(habitat == "eelgrass")

long_kelp <- ba_long %>%
  filter(habitat == "kelp")

eel <- lm(CPUE ~ sp_code + before_after + siteID_NOAA, data = long_eel)
summary(eel)
anova(eel)


kelp <- lm(CPUE ~ sp_code + before_after + siteID_NOAA, data = long_kelp)
summary(kelp)
anova(kelp)
```

# 4. visualize average CPUE by habitat type and species
 Eeelgrass 
```{r}
head(ba_long)

ba_sub <- ba_long %>%
  filter(habitat == "eelgrass", sp_code == "PIPEFISH" | sp_code == "PRICKSN" |
           sp_code == "PERCHSH" | sp_code == "SALCHUM" | sp_code == "STICK3"|
           sp_code == "GUNNCRE" | sp_code == "SCULPSTG" | sp_code == "GREENWHI" |
           sp_code == "GUNNCRE" | sp_code == "SANDLNCP"| sp_code == "SCULNOR" |
           sp_code == "UNARTE" | sp_code == "SCULPAD")

means <- ba_sub %>%
  group_by(sp_code, before_after) %>%
  dplyr::summarise(mean_cpue = mean(CPUE), sd_cpue = sd(CPUE)) %>%
  mutate(before_after = fct_relevel(before_after, "before", "after"))

means$sp_code <- as.character(means$sp_code)

ymin <- means$mean_cpue - means$sd_cpue
ymax <- means$mean_cpue + means$sd_cpue

ggplot(data = means, aes(x = sp_code, y = mean_cpue, fill = before_after)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = ymin, ymax = ymax, x = sp_code),
                position = position_dodge(0.9), width = 0.2) +
  theme_classic() +
  labs(y = "Average CPUE ± s.d.", x = "Species") +
  ggtitle("Eelgrass") +
  guides(fill = guide_legend(title = "Legend")) +
  scale_fill_manual(values = c("#56B4E9", "#E69F00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 24))

```

Kelp
```{r}
means2 <- ba_long %>%
  filter(habitat == "kelp", sp_code == "CAPELIN" | sp_code == "PIPEFISH" |
           sp_code == "SALPINK" | sp_code == "SALCHUM" | sp_code == "GREENWHI" |
           sp_code == "SALCOHO" | sp_code == "SANDLNCP" | sp_code == "TUBESNT" |
           sp_code == "UNARTE" | sp_code == "UNGREEN" | sp_code == "UNROCK") %>%
  group_by(sp_code, before_after) %>%
  dplyr::summarise(mean_cpue = mean(CPUE), sd_cpue = sd(CPUE)) %>%
  mutate(before_after = fct_relevel(before_after, "before", "after"))

ymin <- means2$mean_cpue - means2$sd_cpue
ymax <- means2$mean_cpue + means2$sd_cpue

ggplot(data = means2, aes(x = sp_code, y = mean_cpue, fill = before_after)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = ymin, ymax = ymax, x = sp_code), 
                position = position_dodge(0.9), width = 0.1) +
  theme_classic() +
  labs(y = "Average CPUE ± s.d.", x = "Species") +
  ggtitle("Understory kelp") +
  guides(fill = guide_legend(title = "Legend")) +
  scale_fill_manual(values = c("#56B4E9", "#E69F00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 24))
  
```

## 4.1 Determine if there are any violations of assumptions

```{r}
library("ggpubr")
ggboxplot(ba_sub, x = "sp_code", y = "CPUE", color = "before_after",
          palette = c("#00AFBB", "#E7B800"))


mod1 <- lm(CPUE+1 ~ sp_code + before_after, data = ba_sub)
summary(mod1)

plot(mod1)

e <- residuals(mod1)
y.hat <- fitted.values(mod1)
s <- summary(mod1)$sigma
r <- e/s
d <- rstudent(mod1)

plot(e~y.hat, xlab = "fitted values", ylab = "residuals")

bf.ttest(e, y.hat, cutoff = 50) # unequal variance
shapiro.test(e) # rejects normality

# use boxcox to try another transformation
library(MASS)
boxcox(mod1)
boxcox(mod1, lambda=seq(from=-0.6,to=-.2,by=.1))

# log transform data
mod2 <- lm(log(CPUE +1) ~ sp_code + before_after, data = ba_sub)
plot(mod2)
e <- residuals(mod2)
y.hat <- fitted.values(mod2)

bf.ttest(e, y.hat, cutoff = 2)
shapiro.test(e)

# 4th root transformation
mod3 <- lm((CPUE^(0.25)) ~ sp_code + before_after, data = ba_sub)
plot(mod3)
e <- residuals(mod3)
y.hat <- fitted.values(mod3)

bf.ttest(e, y.hat, cutoff = 2)
shapiro.test(e)

# normality can't really be met...
```

## 4.2 look at glms

```{r}
library(visreg)
mod4 <- glm.nb(CPUE ~ sp_code + before_after, data = ba_sub)

summary(mod4)
plot(mod4)

visreg(mod4)
```

